### 2020-04-30 serial.tex
losing focus

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{diagram.pdf}
\caption{
\textbf{Estimates of the reproduction number from the observed serial intervals.}
(A) Changes in the observed and cohort-averaged mean serial interval over time.
(B) Changes in the estimate of \RR based on the observed and cohort-averaged mean serial interval over time.
Each line represents an independent stochastic realization.
}
\label{fig:obsrR}
\end{figure}

Now, we turn to practical issues.
Serial intervals are typically calculated from line-list data (\fref{obsrR}A).
When an epidemic is ongoing, the observed serial intervals are subject to right-censoring because we cannot observe a serial interval if either an infector or an infectee has not yet developed symptoms;
for example, if we were to measure serial intervals on Day 8 in \fref{obsrR}A, we will only be able to observe the first 6 events (ID 1--6).
\fref{obsrR}B demonstrates how the effect of right-censoring in the observed serial intervals translates to the underestimation of \RR.
Notably, even if we can observe \emph{all} serial intervals across all transmission pairs after the epidemic has ended, we still underestimate the initial mean forward serial interval and therefore \Ro by a large amount because the observed serial-interval distribution converges to the intrinsic serial-interval distribution;
we still underestimate the intrinsic value slightly due to contraction of the forward generation-interval distribution during the susceptible depletion phase (\fref{epi}C).

We provide a simple, heuristic way of assessing potential biases in the estimate of the mean initial forward serial interval and therefore \RR retrospectively.
Once serial intervals have been observed after the epidemic has been sufficiently progressed, we can group observed serial intervals by their primary cohort times (i.e., symptom onset times of infectors).
Then, we can compare how estimates of the mean serial interval as well as \RR change as we include more recent cohorts into the analysis (see `cohort-averaged' in \fref{obsrR}).
During the exponential growth phase, the estimates of the mean serial interval and \RR are consistent with the target value;
adding more data allows us to make more precise inference during this period.
However, the cohort-averaged estimates decreases rapidly soon after the exponential growth period.
This approach allows us to characterize potential biases in the estimates of \RR caused by changes in the forward serial-interval distributions.

@@

Finally, we reanalyze serial intervals of COVID-19 collected by \cite{du2020serial} from mainland China, outside Hubei province, based on transmission events reported between January 21--February 8, 2020;
\cite{du2020serial} estimated the mean serial of 3.96 days (95\% CI 3.53–4.39 days) and \Ro of 1.32 (95\% CI 1.16–1.48).
\fref{du}A shows that the mean forward serial interval decreases over time.
While the decrease is likely to be affected by the right-censoring (indicated by the closeness between the maximum observed serial intervals and maximum \emph{observable} serial intervals),  increase in the proportion of negative serial intervals are clearly indicative of the changes in the forward serial-interval distribution.
\fref{du}B shows that the mean backward serial interval increases over time.
While the qualitative changes in the mean forward and backward serial interval are consistent with our earlier simulations (\fref{epi}), the initial mean forward serial interval (\fref{du}A) appears to be larger than what we calculate earlier based on previously estimated incubation period and generation-interval distributions (\fref{rR}C).
These results indicate that previous estimates of incubation period and generation-intervals may have been underestimated as neither study explicitly accounts for right-censoring (Table 1).
%We do not explicitly account for the right-censoring as it is beyond the scope of this study.

>>>>

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{diagram.pdf}
\caption{
\textbf{Estimating the reproduction number from the observed serial intervals.}
(A) Schematic representation of line list data collected during an epidemic.
(B) Changes in the estimate of \RR based on the observed serial intervals over time.
(C) Schematic representation of line list data rearranged by symptom onset date of infectors.
(B) Changes in the estimate of \RR based on the cohort-averaged serial intervals over time.
}
\label{fig:obsrR}
\end{figure}

Now, we turn to practical issues.
Serial intervals are typically calculated from line-list data (\fref{obsrR}A).
When an epidemic is ongoing, the observed serial intervals are subject to right-censoring because we cannot observe a serial interval if either an infector or an infectee has not yet developed symptoms;
for example, if we were to measure serial intervals on Day 8 in \fref{obsrR}A, we will only be able to observe the first 6 events (ID 1--6).
\fref{obsrR}B demonstrates how the effect of right-censoring in the observed serial intervals translates to the underestimation of \RR in our stochastic simulations.
Notably, even if we can observe \emph{all} serial intervals across all transmission pairs after the epidemic has ended, we still underestimate the initial mean forward serial interval and therefore \Ro by a large amount because the observed serial-interval distribution converges to the intrinsic serial-interval distribution;
we still underestimate the intrinsic value slightly due to contraction of the forward generation-interval distribution during the susceptible depletion phase (\fref{epi}C).

We provide a simple, heuristic way of assessing potential biases in the estimate of the mean initial forward serial interval and therefore \RR retrospectively.
Once serial intervals have been observed after the epidemic has been sufficiently progressed, we can rearrange the line list and group observed serial intervals based on the symptom onset date of infectors (\fref{obsrR}C).
Then, we can compare how estimates of the mean serial interval as well as \RR change as we incorporate more recent cohorts into the analysis;
that is, we analyze observed serial intervals from infectors who became symptomatic before time $t$ and evaluate how the estimates change as we increase $t$
(\fref{obsrR}D).
During the exponential growth phase, the estimates of the mean serial interval and \RR are consistent with the target value;
adding more data allows us to make more precise inference during this period.
However, the cohort-averaged estimates decrease rapidly soon after the exponential growth period, reflecting changes in the forward serial-interval distributions.
This approach allows us to detect dynamical changes in the forward serial-interval distributions and its effect on the estimates of \RR.

@@

Finally, we re-analyze serial intervals of COVID-19 collected by \cite{du2020serial} from mainland China, outside Hubei province, based on transmission events reported between January 21--February 8, 2020;
\cite{du2020serial} estimated the mean serial of 3.96 days (95\% CI 3.53–4.39 days) and \Ro of 1.32 (95\% CI 1.16–1.48).
\fref{du}A shows that the mean forward serial interval decreases over time.
While the decrease is likely to be affected by the right-censoring (indicated by the closeness between the maximum observed serial intervals and maximum \emph{observable} serial intervals),  increase in the proportion of negative serial intervals are clearly indicative of the changes in the forward serial-interval distribution.
\fref{du}B shows that the mean backward serial interval increases over time.
While the qualitative changes in the mean forward and backward serial interval are consistent with our earlier simulations (\fref{epi}), the initial mean forward serial interval (\fref{du}A) appears to be larger than what we calculate earlier based on previously estimated incubation period and generation-interval distributions (\fref{rR}C).
These results indicate that previous estimates of incubation period and generation-intervals may have been underestimated as neither study explicitly accounts for right-censoring (Table 1).
%We do not explicitly account for the right-censoring as it is beyond the scope of this study.
